Unshuffling Permutations

### 2015-07-02 article.tex
Ajouts divers

We denote by $S_n$ the set of permutations of size $n$ and by $S$ the
set of all permutations. If $\pi$ is a permutation, we denote by $|\pi|$
its size. If $P$ is a subset of $[|\pi|]$, $\pi_{|P}$ is the subword of
$\pi$ consisting in the letters of $\pi$ at the positions specified by
the elements of $P$. Besides, we denote by $\widetilde{\pi}$ (resp.
$\bar \pi$, $\pi^{-1}$) the mirror image (resp. complementary, inverse)
of $\pi$. If $u$ is a word of integers without multiple occurrences of a
same letter, $\STD(u)$ denotes the unique permutation of same size as
$u$ such that for all $i, j \in [|u|]$, $u_i < u_j$ if and only if
$\STD(u)_i < \STD(u)_j$.
\medskip

@@

%%
%% ---- Algebraic issues ----
%%
\section{Algebraic issues}
\label{section:Algebraic issues}
Let use denote by $\QQ[S]$ the linear span of all permutations. We shall
define on this space a kind of shuffle product. For this purpose, we
first define a linear coproduct $\Delta$ on $\QQ[S]$ satisfying for
any permutation $\pi$,
\begin{equation} \label{equ:unshuffling_coproduct}
    \Delta(\pi) =
    \sum_{P_1 \uplus P_2 = [|\pi|]}
    \STD\left(\pi_{|P_1}\right) \otimes \STD\left(\pi_{|P_2}\right).
\end{equation}
We call $\Delta$ the {\em unshuffling coproduct of permutations}. For
instance,
\begin{equation}
    \Delta(213) =
    \epsilon \otimes 213 + 2. 1 \otimes 12 +
    1 \otimes 21 + 2. 12 \otimes 1 + 21 \otimes 1 + 213 \otimes \epsilon,
\end{equation}
\begin{equation}
    \Delta(1234) =
    \epsilon \otimes 1234 + 4. 1 \otimes 123 + 6. 12 \otimes 12 +
    4. 123 \otimes 1 + 1234 \otimes \epsilon,
\end{equation}
\begin{equation}\begin{split} \label{equ:example_unshuffling_coproduct}
    \Delta(1432) & =
    \epsilon \otimes 1432 + 3. 1 \otimes 132 + 1 \otimes 321 +
    3. 12 \otimes 21 \\ & + 3. 21 \otimes 12 + 3. 132 \otimes 1 +
    321 \otimes 1 + 1432 \otimes \epsilon.
\end{split}\end{equation}
Observe that the coefficient of the tensor $1 \otimes 132$ is $3$
in~\eqref{equ:example_unshuffling_coproduct} because there are exactly
three ways to extract two disjoint subwords of the permutation $1432$
so that the letters of these two subwords have the same relative order
as the letters of the permutations $1$ and $132$.
\medskip

@@

    If $P$ is a set of indexes of letters of $\pi$, we denote by
    $\widetilde{\pi}$ the set $\{|\pi| - i  + 1 : i \in P\}$. Now since
    this operation $\widetilde{\,}$ define a bijection on the set of the
    subsets of $[|\pi|]$ and the standardisation operation commutes with
    the mirror operation on words, we have
    \begin{equation} \begin{split}
        \Delta(\phi_1(\pi))
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\phi_1(\pi)_{|P_1}\right)
        \otimes \STD\left(\phi_1(\pi)_{|P_2}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\widetilde{\pi}_{|P_1}\right)
        \otimes \STD\left(\widetilde{\pi}_{|P_2}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\widetilde{\pi}_{|\widetilde{P_1}}\right)
        \otimes \STD\left(\widetilde{\pi}_{|\widetilde{P_2}}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \widetilde{\STD\left(\pi_{|P_1}\right)}
        \otimes \widetilde{\STD\left(\pi_{|P_2}\right)} \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \phi_1\left(\STD\left(\pi_{|P_1}\right)\right)
        \otimes \phi_1\left(\STD\left(\pi_{|P_2}\right)\right) \\
        & = (\phi_1 \otimes \phi_1) \Delta(\pi).
    \end{split} \end{equation}
    This shows that $\phi_1$ is a morphism of coalgebras and hence, that
    $\phi_1$ is a morphism of associative algebras.
    \smallskip

@@

    Finally, for any permutation $\tau$, if $P$ is a set of indexes of
    letters of $\tau$, we denote by $P_\tau^{-1}$ the set
    $\{\tau_i : i \in P\}$. Since the map sending a subset $P$ of
    $[|\pi|]$ to $P_\pi^{-1}$ is a bijection and
    $\STD\left(\pi_{|P}\right)^{-1} = \STD\left(\pi^{-1}_{|P_\pi^{-1}}\right)$,
    we have
    \begin{equation} \begin{split}
        \Delta(\phi_3(\pi))
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\phi_3(\pi)_{|P_1}\right)
        \otimes \STD\left(\phi_3(\pi)_{|P_2}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\pi^{-1}_{|P_1}\right)
        \otimes \STD\left(\pi^{-1}_{|P_2}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\pi^{-1}_{|{P_1}_\pi^{-1}}\right)
        \otimes \STD\left(\pi^{-1}_{|{P_2}_\pi^{-1}}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\pi_{|P_1}\right)^{-1}
        \otimes \STD\left(\pi_{|P_2}\right)^{-1} \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \phi_3\left(\STD\left(\pi_{|P_1}\right)\right)
        \otimes \phi_3\left(\STD\left(\pi_{|P_2}\right)\right) \\
        & = (\phi_3 \otimes \phi_3) \Delta(\pi).
    \end{split} \end{equation}
    This shows that $\phi_3$ is a morphism of coalgebras and hence, that
    $\phi_3$ is a morphism of associative algebras.
\end{proof}
\medskip

>>>>

We denote by $S_n$ the set of permutations of size $n$ and by $S$ the
set of all permutations. If $\pi$ is a permutation, we denote by $|\pi|$
its size. The unique permutation of size $0$ is denoted by $\epsilon$.
If $P$ is a subset of $[|\pi|]$, $\pi_{|P}$ is the subword of $\pi$
consisting in the letters of $\pi$ at the positions specified by the
elements of $P$. Besides, we denote by $\widetilde{\pi}$ (resp.
$\bar \pi$, $\pi^{-1}$) the mirror image (resp. complementary, inverse)
of $\pi$. If $u$ is a word of integers without multiple occurrences of a
same letter, $\STD(u)$ denotes the unique permutation of same size as
$u$ such that for all $i, j \in [|u|]$, $u_i < u_j$ if and only if
$\STD(u)_i < \STD(u)_j$.
\medskip

@@

%%
%% ---- Algebraic issues ----
%%
\section{Algebraic issues}
\label{section:Algebraic issues}
The purpose of this section is to define a shuffle product $\SHUFFLE$
on permutations and establish some of its properties. To define this
shuffle product, we shall first define a coproduct~\cite{Joni:Rota:1979}
denoted by $\Delta$. By duality, $\Delta$ implies the definition of
$\SHUFFLE$. It is worth to note that, as we will see, algebraic
properties of $\Delta$ imply combinatorial properties of $\SHUFFLE$.
\medskip

Let use denote by $\QQ[S]$ the linear span of all permutations. As
announced, we shall define on this space a kind of shuffle product and
this purpose, we first define a linear coproduct $\Delta$ on $\QQ[S]$.
For any permutation $\pi$, we set
\begin{equation} \label{equ:unshuffling_coproduct}
    \Delta(\pi) =
    \sum_{P_1 \uplus P_2 = [|\pi|]}
    \STD\left(\pi_{|P_1}\right) \otimes \STD\left(\pi_{|P_2}\right).
\end{equation}
We call $\Delta$ the {\em unshuffling coproduct of permutations}. For
instance,
\begin{equation}
    \Delta(213) =
    \epsilon \otimes 213 + 2. 1 \otimes 12 +
    1 \otimes 21 + 2. 12 \otimes 1 + 21 \otimes 1 + 213 \otimes \epsilon,
\end{equation}
\begin{equation}
    \Delta(1234) =
    \epsilon \otimes 1234 + 4. 1 \otimes 123 + 6. 12 \otimes 12 +
    4. 123 \otimes 1 + 1234 \otimes \epsilon,
\end{equation}
\begin{equation}\begin{split} \label{equ:example_unshuffling_coproduct}
    \Delta(1432) & =
    \epsilon \otimes 1432 + 3. 1 \otimes 132 + 1 \otimes 321 +
    3. 12 \otimes 21 \\ & + 3. 21 \otimes 12 + 3. 132 \otimes 1 +
    321 \otimes 1 + 1432 \otimes \epsilon.
\end{split}\end{equation}
Observe that the coefficient of the tensor $1 \otimes 132$ is $3$
in~\eqref{equ:example_unshuffling_coproduct} because there are exactly
three ways to extract two disjoint subwords of the permutation $1432$
so that the letters of these two subwords have the same relative order
as the letters of the permutations $1$ and $132$.
\medskip

@@

    If $P$ is a set of indexes of letters of $\pi$, we denote by
    $\widetilde{\pi}$ the set $\{|\pi| - i  + 1 : i \in P\}$. Now, since
    this operation $\widetilde{\,}$ define a bijection on the set of the
    subsets of $[|\pi|]$, and since the standardization operation
    commutes with the mirror operation on words, we have
    \begin{equation} \begin{split}
        \Delta(\phi_1(\pi))
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\phi_1(\pi)_{|P_1}\right)
        \otimes \STD\left(\phi_1(\pi)_{|P_2}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\widetilde{\pi}_{|P_1}\right)
        \otimes \STD\left(\widetilde{\pi}_{|P_2}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\widetilde{\pi}_{|\widetilde{P_1}}\right)
        \otimes \STD\left(\widetilde{\pi}_{|\widetilde{P_2}}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \widetilde{\STD\left(\pi_{|P_1}\right)}
        \otimes \widetilde{\STD\left(\pi_{|P_2}\right)} \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \phi_1\left(\STD\left(\pi_{|P_1}\right)\right)
        \otimes \phi_1\left(\STD\left(\pi_{|P_2}\right)\right) \\
        & = (\phi_1 \otimes \phi_1) \Delta(\pi).
    \end{split} \end{equation}
    This shows that $\phi_1$ is a morphism of coalgebras and hence, that
    $\phi_1$ is a morphism of associative algebras.
    \smallskip

@@

    Finally, for any permutation $\tau$, if $P$ is a set of indexes of
    letters of $\tau$, we denote by $P_\tau^{-1}$ the set
    $\{\tau_i : i \in P\}$. Since the map sending a subset $P$ of
    $[|\pi|]$ to $P_\pi^{-1}$ is a bijection, and since
    $\STD\left(\pi_{|P}\right)^{-1} = \STD\left(\pi^{-1}_{|P_\pi^{-1}}\right)$,
    we have
    \begin{equation} \begin{split}
        \Delta(\phi_3(\pi))
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\phi_3(\pi)_{|P_1}\right)
        \otimes \STD\left(\phi_3(\pi)_{|P_2}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\pi^{-1}_{|P_1}\right)
        \otimes \STD\left(\pi^{-1}_{|P_2}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\pi^{-1}_{|{P_1}_\pi^{-1}}\right)
        \otimes \STD\left(\pi^{-1}_{|{P_2}_\pi^{-1}}\right) \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \STD\left(\pi_{|P_1}\right)^{-1}
        \otimes \STD\left(\pi_{|P_2}\right)^{-1} \\
        & = \sum_{P_1 \uplus P_2 = [|\pi|]}
        \phi_3\left(\STD\left(\pi_{|P_1}\right)\right)
        \otimes \phi_3\left(\STD\left(\pi_{|P_2}\right)\right) \\
        & = (\phi_3 \otimes \phi_3) \Delta(\pi).
    \end{split} \end{equation}
    This shows that $\phi_3$ is a morphism of coalgebras and hence, that
    $\phi_3$ is a morphism of associative algebras.
\end{proof}
\medskip

We now use the algebraic properties of $\SHUFFLE$ exhibited by
Proposition~\ref{prop:endomorphisms} to obtain combinatorial properties
of square permutations.
\medskip
